Average solar wind and geomagnetic activity influence on the equatorial plasma mass density at geosynchronous orbit

### 2016-08-04 paper/paper.tex
Adding MLT cutoff to Fig 4
Fixing the issues the git merge caused
Updated abstract and results

%\twocolumn[
%  \begin{@twocolumnfalse}
%\maketitle
%\hrule
\begin{abstract}
We consider two types of events, identified by decreases in $D_{st}$ below a threshold value and increases in the the equatorial mass density at geosynchronous altitudes, $\rho_{eq}$, above a threshold value on 1-hour and 1-day time scales using the \cite{Takahashi2010} dataset.  From the $D_{st}$ events, we find that there is a weak and small amplitude ($< 10$\%) difference between $\rho_{eq}$ on the day of the event and the days before and after on both daily and hourly timescales.  Although very large geomagnetic storms have been observed to correspond to large increases in $\rho_{eq}$, on average, $\rho_{eq}$ following the onset of a geomagnetic storm does not depend on the north-south component of the interplanetary magnetic field, $B_z$ near onset time. From the $\rho_{eq}$ events, we find a weak dependence on $B_{z}$ after the onset of an event, with higher average $B_{z}$ four hours after the event onset corresponding to larger $\rho_{eq}$ events.  In contrast, the average of $B_z$ four hours prior to $\rho_{eq}$ events does not have a statistically significant impact on $\rho_{eq}$ after onset.
\end{abstract}
%\vspace{2em}
%\hrule
%\vspace{2em}
% \end{@twocolumnfalse}
%  ]

@@

Our first analysis uses GOES-6 data only from the time interval 1989-1991, which corresponds with the interval used in the epoch analysis of \cite{Takahashi2010}. In this interval, there were 75 $D_{st}$ events where $D_{st}$ remained below the $-50$~nT threshold for at least 12 hours.  For both types of events, the first hour of a local minimum $D_{st}$ following a threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before and $8\cdot24$ hours after event onset. For multiple detected threshold crossings within 24 hours of each other, the first crossing was selected as the event since visual inspection showed the following crossings to typically be from brief deviations during the recovery period. To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians for each event in 1-day bins and then compute the medians across events on each epoch day.

@@

If a statistically significant difference in the medians between two days existed, the fraction of bootstrap samples with a different sign is expected to be much smaller ($<$ 5\%).  In terms of a hypothesis test, we cannot reject the null hypothesis that the medians are the same with a confidence higher than 100 minus the \% value shown in the table; the highest confidence level of 89\% is in the difference between day 0 and day -1.

These tests indicate that there is not a statistically significant difference in medians of $\rho_{eq}$ on the day before and day after a threshold crossing in $D_{st}$.  In addition, the observed differences are on the order of only 20\% of the value of $\rho_{eq}$ on the onset day, which is similar to the variation in the observations over the displayed epoch time interval.

@@

We have also considered hourly averages as shown in the top panel of Figure~\ref{HourlyAveragedDstEvents}.  A total of 333 $D_{st}$ events were found in the interval of May 1983 through August 1991 with an average duration of 10 hours and a median duration of 3 hours. In this plot, the overall trend is similar to that found in the bottom panel of Figure~\ref{HourlyAveragedDstEvents} - $\rho_{eq}$ is slightly elevated at the time of the threshold crossing of $D_{st}$ with respect to the values one day before and after, but the elevation is small relative to the error bars around the time of the event onset.

The bottom panel of Figure~\ref{HourlyAveragedDstEvents} contains the events shown in the top panel under the constraint that $D_{st}$ remained below -50~nT for at least twelve hours after the event onset time.  In the time period of elevated mass density within 12 hours of the start of the event, there are only $\sim$4 samples per hourly interval.  This result is consistent with previous observations of large mass density increases in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010}).

The top panel of Figure~\ref{HourlyAveragedDstEvents} shows that, on average, near the onset of $D_{st}$ events there is not a significant reponse in $\rho_{eq}$ while the bottom panel shows that for rare and large $D_{st}$ events elevated $\rho_{eq}$ can be observed.  To determine if there is an IMF $B_{z}$ dependence, the epoch curves in the top panel of Figure~\ref{HourlyAveragedDstEvents} for $\rho_{eq}$ and $B_{z}$ were separated into two parts depending whether $B_z$ was above or below its average four hours before and including the onset hour and four hours after and including the hour of onset; the result is shown in Figure~\ref{fig:BzBinnedDstEvents}. Any statistically significant difference at a 95\% confidence level between the two parts is indicated by a green dot on the horizontal axis using the same bootstrap method described previously.  (At this confidence level, false positives are expected to randomly occur 5\% of the time, or 3.6 times for 72 values.)  These two curves indicate that there is not a statistically significant dependence on IMF $B_{z}$ around the time of the onset of a $D_{st}$ event.

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{figures/stormavs-dst-GOES6.eps}
\\
\rule[1ex]{5cm}{1pt}
\\
\includegraphics[scale=0.40]{figures/stormavs-dd12-GOES6.eps}
\caption{$D_{st}$ events from GOES 6 using hourly medians. Top: Events in the interval 1983-1991. Bottom: Same as Top except for constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours after crossing below $-50$~nT.}
\label{HourlyAveragedDstEvents}
\end{figure}

>>>>

%\twocolumn[
%  \begin{@twocolumnfalse}
%\maketitle
%\hrule
\begin{abstract}
We consider two types of events, identified by decreases in $D_{st}$ below a threshold value and increases in the the equatorial mass density at geosynchronous altitudes, $\rho_{eq}$, above a threshold value on 1-hour and 1-day time scales using the \cite{Takahashi2010} dataset.  From the $D_{st}$ events, we find that there is a statistically weak and small amplitude difference between $\rho_{eq}$ on the day of the event and the days before and after on daily timescales.  On hourly timescales, it is observed that for 6 hours after onset, very few measurements of  $\rho_{eq}$ are found and that most of the contribution to the daily averages is from 6-24 hours after onset.  On hourly timescales, $\rho_{eq}$ following the onset of a $D_{st}$ event does not depend on the north-south component of the interplanetary magnetic field, $B_z$ near onset time. The primary factor that determines if a post-$D_{st}$ event onset peak in $\rho_{eq}$ will occur is elevated $F10.7$.  From the $\rho_{eq}$ events, we find a weak dependence on $B_{z}$ after the onset of an event, with higher average $B_{z}$ four hours after the event onset corresponding to larger $\rho_{eq}$ values 24-36 hours after onset.  In contrast, the average of $B_z$ four hours prior to $\rho_{eq}$ events does not have a statistically significant impact on $\rho_{eq}$ after onset.
\end{abstract}
%\vspace{2em}
%\hrule
%\vspace{2em}
% \end{@twocolumnfalse}
%  ]

@@

Our first analysis uses GOES-6 data only from the time interval 1989-1991, which corresponds with the interval used in the epoch analysis of \cite{Takahashi2010}. In this interval, there were 75 $D_{st}$ events where $D_{st}$ remained below the $-50$~nT threshold for at least 12 hours. Only events with an onset time between 06 and noon MLT were considered. For both types of events, the first hour of a local minimum in $D_{st}$ following a threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before and $8\cdot24$ hours after event onset. For multiple detected threshold crossings within 24 hours of each other, the first crossing was selected as the event since visual inspection showed the following crossings to typically be from brief deviations during the recovery period of a geomagnetic storm. To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians for each event in 1-day bins and then compute the medians across events on each epoch day.

@@

If a statistically significant difference in the medians between two days existed, the fraction of bootstrap samples with a different sign is expected to be much smaller ($<$ 5\%).  In terms of a hypothesis test, we cannot reject the null hypothesis that the medians are the same with a confidence higher than 100 minus the \% value shown in the table; the highest confidence level of 86\% is in the difference between day 0 and day -1.

A similar hypothesis test run for the bottom panel of Figure~\ref{DailyAveragedDstEvents} yields a maximum confidence level of $\sim$85\%.

These tests indicate that there is not a statistically significant difference in medians of daily averaged $\rho_{eq}$ on the day before and day after a threshold crossing in $D_{st}$.  In addition, the observed differences are on the order of only 20\% of the value of $\rho_{eq}$ on the onset day, which is similar to the variation in the observations over the displayed epoch time interval.

@@

We have also considered the hourly averages from the events shown in the bottom panel of Figure~\ref{HourlyAveragedDstEvents}, which shown in the top panel of Figure~\ref{HourlyAveragedDstEvents}.  In the time period of elevated mass density within 6 hours of the start of the event, there are only $\sim$4 samples per hourly interval.  The peak value of $\sim$70 amu/cm$^3$ is associated with a single observation.

To increase the number events, we have remove the constraint that the events must occur in the MLT range from 06-12 and the constraint that $D_{st}$ remained below the threshold value for 12 hours or longer.  The result is shown in the bottom panel of Figure~\ref{HourlyAveragedDstEvents}.

%This result is consistent with previous observations of large mass density increases in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010}).

%A total of 333 $D_{st}$ events were found in the interval of May 1983 through August 1991 with an average duration of 10 hours and a median duration of 3 hours. In this plot, the overall trend is similar to that found in the bottom panel of Figure~\ref{HourlyAveragedDstEvents} - $\rho_{eq}$ is slightly elevated at the time of the threshold crossing of $D_{st}$ with respect to the values one day before and after, but the elevation is small relative to the error bars around the time of the event onset.

%The top panel of Figure~\ref{HourlyAveragedDstEvents} shows that, on average, near the onset of $D_{st}$ events there is not a significant reponse in $\rho_{eq}$ while the bottom panel shows that for rare and large $D_{st}$ events elevated $\rho_{eq}$ can be observed.

To determine if there is an IMF $B_{z}$ dependence, the epoch curves in the top panel of Figure~\ref{HourlyAveragedDstEvents} for $\rho_{eq}$ and $B_{z}$ were separated into two parts depending whether $B_z$ was above or below its average four hours before and including the onset hour and four hours after and including the hour of onset; the result is shown in Figure~\ref{fig:BzBinnedDstEvents}. Any statistically significant difference at a 95\% confidence level between the two parts is indicated by a green dot on the horizontal axis using the same bootstrap method described previously.  (At this confidence level, false positives are expected to randomly occur 5\% of the time, or 3.6 times for 72 values.)  These two curves indicate that there is not a statistically significant dependence on IMF $B_{z}$ around the time of the onset of a $D_{st}$ event.

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{figures/stormavs-dd12-GOES6.eps}
\\
\rule[1ex]{5cm}{1pt}
\\
\includegraphics[scale=0.40]{figures/stormavs-dst-GOES6.eps}
\caption{$D_{st}$ events from GOES 6 using hourly medians. Top: Events with the constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours after crossing below $-50$~nT.  Bottom: Same as Top except without constraint.}
\label{HourlyAveragedDstEvents}
\end{figure}
